---
output:
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# bdfm
Bayesian dynamic factor model estimation in R.

[![Build Status](https://travis-ci.org/christophsax/tsbox.svg?branch=master)](https://travis-ci.org/srlanalytics/bdfm)
[![codecov](https://codecov.io/github/srlanalytics/bdfm/branch/master/graphs/badge.svg)](https://codecov.io/github/srlanalytics/bdfm)

bdfm estimates dynamic factor models by simulation using the Durbin and Koopman
(2012) disturbance smoother and returns estimated factors, predicted values of
observables, posterior distributions for predicted values (optional), and
forecast updates by series (optional). Maximum likelihood estimation via Watson
and Engle (1983) is also supported. Input data may be noisy, have missing
values, or "ragged edges" due to different start or end dates.

## Installation

To install the development version:

```r
remotes::install_github("srlanalytics/bdfm")
```

## Basic Use

Estimate a model with one factor and two lags in the transition equation:
```{r, message = FALSE}
library(bdfm)
dta <- cbind(fdeaths, mdeaths)
m <- dfm(dta, forecast = 2)
```

Input data should be an object of class `"ts"`. If the
[tsbox](https://www.tsbox.help/) package is installed, any ts-boxable time
series objects can be supplied, such as `"xts"`, `"zoo"`, `"timeSeries"` or
`"data.frame"`.

Extract the resulting series, using `predict()`.

```{r predict}
plot(predict(m))
```


Or have a look at the esimated factors, using `factors()`:


```{r factors}
plot(factors(m))
```


## Estimation Details

You can find details on estimation routines and derivations in the short book
*Practical Implementation of Factor Models*. [Free
Download](http://srlquantitative.com)


## Related Packages

[MARSS](https://cran.r-project.org/web/packages/MARSS/index.html), [dynfactoR](https://rdrr.io/github/guilbran/dynfactoR/)
: Offer maximum likelihood estimation only

[greta](https://cran.r-project.org/web/packages/greta/index.html)
: Uses Hamiltonian Monte Carlo simulation find posterior distributions of model parameters for whatever model the user specifies. It looks superb, but being general purpose it is much more difficult to implement DFMs than with **bdfm**.

